DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-06-06 manuscript/figures-alt.tex
formatting legends and adding png figure for fringillidae example
addressing remaining comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
Bird species within the Fringillidae are paraphyletic
(Alström et al. \protect\hyperlink{ref-}{2014},
Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
Price et al. \protect\hyperlink{ref-price2014niche}{2014},
Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
}
\label{fig:fringillidae-topologies}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
\label{fig:fringillidages}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%    FIGURE TOPOLOGIES    %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
Bird species within the Fringillidae are paraphyletic
(Alström et al. \protect\hyperlink{ref-}{2014},
Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
Price et al. \protect\hyperlink{ref-price2014niche}{2014},
Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
}
\label{fig:fringillidae-topologies}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%    FIGURE FRINGILLIDAE    %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
\label{fig:fringillidages}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
\caption{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
\label{fig:cvXY}
\end{figure}

### 2022-06-06 manuscript/figures-supplementary.tex
cross validation XY figure
addressing remaining comments

% in this file I leave the figure captions outside\ caption{} because I want them
% to be formatted in the same way as the general text (double spaced and linenumbered)
\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
\caption{Results from cross validation analysis.}
\label{fig:cvXY}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
\caption{Results from cross validation analysis.}
\label{fig:cvXYdiffs}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
\caption{Cross validation of second source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv2}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
\caption{Cross validation of third source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv3}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
\caption{Cross validation of fourth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv4}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
\caption{Cross validation of sixth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv6}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
\caption{Cross validation of seventh source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv7}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
\caption{Cross validation of eight source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv8}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
\caption{Cross validation of ninth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv9}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
\caption{Cross validation of tenth source chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv10}
\end{figure}

>>>>

% in this file I leave the figure captions outside\ caption{} because I want them
% to be formatted in the same way as the general text (double spaced and linenumbered)
\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Supplementary Figure}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
\caption{Results from cross validation analysis. Each plot shows the difference between the original age estimated and those obtained with a DateLife analysis, per node.}
\label{fig:cvXYdiffs}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%   CROSS VALIDATION CHRONOGRAMS   %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
\caption{Cross validation of Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}) chronogram 1. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv2}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
\caption{Cross validation of Barker et al. (\protect\hyperlink{ref-barker2015new}{2015}) chronogram 1. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv3}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
\caption{Cross validation of Burns et al. (\protect\hyperlink{ref-burns2014phylogenetics}{2014}) chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv4}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
\caption{Cross validation of Gibb et al. (\protect\hyperlink{ref-gibb2015new}{2015}), chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv6}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
\caption{Cross validation of Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}) chronogram 1. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv7}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
\caption{Cross validation of Hedges et al. (\protect\hyperlink{ref-Hedges2015}{2015}) chronogram 2. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv8}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
\caption{Cross validation of Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}) chronogram. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv9}
\end{figure}

\begin{figure}[!h]
\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
\caption{Cross validation of Jetz et al. (\protect\hyperlink{ref-Jetz2012}{2012}) chronogram 1. The chronogram shown in black corresponds to the dates published in the original study. The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
\label{fig:cv10}
\end{figure}

### 2022-06-06 manuscript/manuscript.Rmd
addressing remaining comments

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : ["tables.tex",
                     "figures-alt.tex", 
                     "figures-supplementary.tex"]

@@

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021]. For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
<!--\textcolor{red}{EJM: I don't understand what this sentence means}
LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
LLSR: I moved it here, it makes more sense here, I think!.
-->

@@

_**Searching the database.-**_
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species. 
The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.

_**Summarizing search results.-**_
The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. 
As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology. 
According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).
Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@sauquet2013practical], as for example done by [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases [@ksepka2015fossil], though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->

\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}

It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical]. 
Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses. It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.

@@

Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). This study inferred dates using molecular substitution rate estimates across birds [@weir2008calibrating]
\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}

@@

The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram, 
that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
LLSR: I changed loosing to blurring. But I will elaborate more on it}
Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
\textcolor{red}{LLSR: Example}

<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age stimates across studies is a widespread phenomenon that requires further attention. -->

>>>>

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : ["tables.tex",
                     "figures-alt.tex"]

@@

Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021). For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
<!--\textcolor{red}{EJM: I don't understand what this sentence means}
LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
LLSR: I moved it here, it makes more sense here, I think!.
-->

@@

_**Searching the database.-**_
Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species. 
The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).

_**Summarizing search results.-**_
The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. 
As explained in the "Description" section, a tree topology to summarize age data upon is required.
By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query. 
According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig. \ref{fig:fringillidages}).
Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
<!-- ????????????????? Say some things about the results!
ANSWER:
(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

@@

@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart. <!-- Will this be implemented in future `datelife` versions? Probably not. -->
There have been attempts to create fossil calibration databases [@ksepka2015fossil], though these still have room to grow. <!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves? the age difference between primary and secondary calibrations?}-->

It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical]. 
Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses. It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.

Our cross validation analysis might provide some insight in this regard. When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations. However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs. 1 and 4).

@@

Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs. fig:fringillidages and fig:cvXY; supplementary figs. 1 and 4).
<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

@@

The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram, 
that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS! This is great news. They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
<!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
LLSR: I changed loosing to blurring. But I will elaborate more on it}
Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
\textcolor{red}{LLSR: Example}-->

<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age estimates across studies is a widespread phenomenon that requires further attention. -->
